Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-05-28 docs/_tex/manuscript.tex
add query rate

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data. In total,
the entire study cost \$296.27 in queries to the OpenAI API. While we do
not have a direct comparison, this likely represents a extremely small
fraction of the total costs of such a study if performed manually by
human experts charging at an hourly rate. Even if all human annotations
were provided on a volunteer basis, this would still require hundreds if
not thousands of hours of cumulative manual human labour. Using our
approach, severity annotations for the entire HPO can be generated
automatically in a matter of several hours.

>>>>

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data. In total,
the entire study cost \$296.27 in queries to the OpenAI API. While we do
not have a direct comparison, this likely represents a extremely small
fraction of the total costs of such a study if performed manually by
human experts charging at an hourly rate. Even if all human annotations
were provided on a volunteer basis, this would still require hundreds if
not thousands of hours of cumulative manual human labour. Using our
approach, severity annotations for the entire HPO can be generated
automatically at a rate of \textasciitilde100 phenotypes/hour. Further
optimisation of the annotation process and increased API rate limits
could potentially accelerate this even further.

### 2024-05-28 manuscript.qmd
add query rate

```{r gpt_annot_class, results='hide'}
res_class <- HPOExplorer::gpt_annot_class(annot_coded = res_coded$annot_coded)
res_class <- merge(res_class,
                   res_coded$annot_weighted[,c("hpo_name","severity_score_gpt")])
fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
                            x="severity_class", y="severity_score_gpt")
fig_severity_class_dt <- MSTExplorer:::get_ggstatsplot_stats(fig_severity_class)
```

@@

## Abstract

The Human Phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised framework for describing phenotypic abnormalities. However, documentation on the severity of each phenotype is currently very sparse, making systematic analyses and prioritisation of results challenging. With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO. First, we benchmarked the generative LLM annotations against ground-truth labels within the HPO (e.g. phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4). True positive recall rates across different annotations ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence. These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.

@@

```{r severity_scoring}
## code adapted from HPOExplorer::gpt_annot_plot()
# get all descendants of the term phenotypic abnormality 
keep_ont_levels=seq(3,17)
#### Filter out ont levels ####
annot_melt <- HPOExplorer::add_ancestor(annot_melt,
                                        keep_descendants = keep_descendants)
data.table::setorderv(annot_melt,"severity_score_gpt",-1)

annot_melt <- HPOExplorer::add_ont_lvl(annot_melt, keep_ont_levels = keep_ont_levels)
annot_melt$variable <- gsub("_", " ", annot_melt$variable)

# get all unique phenotypes
total_unique <- length(unique(annot_melt$hpo_id))

# get the top N phenotypes
top_n <- 15
dat_top <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(top_n)]]

# Get the bottom N phenotypes
bottom_n <- 15
dat_bottom <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(total_unique - bottom_n + 1, total_unique)]]

# Define the range for selecting phenotypes from the middle
start_index <- ceiling(total_unique / 2)  
end_index <- start_index + 12

# Get phenotypes from the middle
dat_middle <- annot_melt[hpo_id %in% unique(annot_melt$hpo_id)[seq(start_index, end_index)]]

# define labels
dat_top$position <- "Top"
dat_middle$position <- "Middle"
dat_bottom$position <- "Bottom"

dat_all <- rbind(dat_top, dat_middle, dat_bottom)
dat_all$position <- factor(dat_all$position, levels=unique(dat_all$position))
dat_all$variable <- factor(dat_all$variable, levels = annotation_order)
```

```{r fig-top-phenos}
#| label: fig-top-phenos
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of the top (n = 15), mid-range (n = 15), and bottom (n = 15) HPO phenotypes based on severity score. Severity scores were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights. The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axis is scaled by the severity of the clinical characterisation.
#| fig-height: 7.5
#| fig-width: 8

p1 <- ggplot2::ggplot(data = dat_all,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile(colour = "white", lwd = 0.5, linetype =1) +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_brewer(palette = "GnBu", direction = -1) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right') +
  labs(x = "Clinical characteristic",
       y = "HPO phenotype",
       fill = "Occurrence") +
  ggplot2::facet_grid(position ~ ., scales = "free_y")

p2 <- ggplot2::ggplot(data = dat_all,
                      ggplot2::aes(x="severity_score_gpt", y=hpo_name,
                                   fill=severity_score_gpt)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
  ggplot2::labs(x=NULL, y=NULL, fill = "Severity\nscore") +
  ggplot2::theme_void() +
  guides(fill = guide_colorbar(reverse = TRUE)) +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                 axis.ticks.y = ggplot2::element_blank(),
                 legend.position = 'right')

p3 <- patchwork::wrap_plots(p1, p2, ncol = 2,
                            widths = c(1,.2),
                            guides = "collect")

methods::show(p3)
```

@@

Here, we introduce a novel framework to leverage the current best-in-class LLM, GPT-4 [@openaiGPT4TechnicalReport2024], to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO. By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without substantially compromising accuracy. Our validation approach yielded a high true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across the phenotypes tested. Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data.

Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities. An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt`), which GPT-4 annotated as causing intellectual disability. At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition. However, upon further inspection we realised that molar tooth sign is in fact a pattern of abnormal brain morphology that happens to bear some resemblance to molar dentition when observed in radiological scans. This phenotype is a known sign of neurodevelopmental defects that can indeed cause severe intellectual disability [@gleesonMolarToothSign2004].

>>>>

```{r gpt_annot_class, results='hide'}
res_class <- HPOExplorer::gpt_annot_class(res_coded = res_coded)
fig_severity_class <- ggstatsplot::ggbetweenstats(res_class,
                            x="severity_class", y="severity_score_gpt")
fig_severity_class_dt <- MSTExplorer:::get_ggstatsplot_stats(fig_severity_class)
```

@@

```{r weights_dict}
weights_dict <- eval(formals(gpt_annot_codify)$weights_dict)
annotation_order <- gsub("_"," ",names(sort(unlist(weights_dict), decreasing = TRUE)))
```

## Abstract
There are thousands of human phenotypes which are linked to genetic variation. These range from the benign (white eyelashes) to the deadly (respiratory failure). The Human Phenotype Ontology has categorised all human phenotypic variation into an unified framework that defines the relationships between them (e.g. missing arms and missing legs are both abnormalities of the limb). This has made it possible to perform phenome-wide analyses, e.g. to prioritise which make the best candidates for gene therapies. However, there is currently limited metadata describing the effects / severity of these phenotypes. With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO. First, we benchmarked the generative LLM annotations against ground-truth labels within the HPO (e.g. phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4). True positive recall rates across different annotations ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence. These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.

@@

```{r fig-top-phenos}
#| label: fig-top-phenos
#| fig-cap: Quantifying the severity of HPO phenotype annotations highlights the most impactful conditions. Heatmap of 10 represetantive phenotypes from each severity class (Profound, Severe, Moderate, Mild) stratified by whether the phenotypes are often/always congenital (**a**-**b**) or rarely/never congenital (**c**-**d**). Continuous severity scores are shown as bars (**b**,**d**) and were calculated by multiplying the numeric values assigned to each clinical characteristic according to @tbl-metric-weights. The average normalised score, representing overall phenotype severity on a 0-100 scale, was calculated by aggregating the multiplied values and normalising by the theoretical maximum severity score. The x-axes show each of the clinical characteristics. All data for this figure, as well as justifications for each annotation, can be found in @tbl-annotations.
#| fig-height: 10
#| fig-width: 13

plot_top_phenos_out <- HPOExplorer::plot_top_phenos(res_class = res_class)
plot_top_phenos_out$plot
```

@@

Here, we introduce a novel framework to leverage the current best-in-class LLM, GPT-4 [@openaiGPT4TechnicalReport2024], to systematically annotate the severity of `r n_ids_gpt` phenotypic abnormalities within the HPO. By employing advanced AI capabilities, we have demonstrated the feasibility of automating this process, significantly enhancing efficiency without substantially compromising accuracy. Our validation approach yielded a high true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across the phenotypes tested. Furthermore, our approach can be readily adapted and scaled to accommodate the growing volume of phenotypic data. In total, the entire study cost $296.27 in queries to the OpenAI API. While we do not have a direct comparison, this likely represents a extremely small fraction of the total costs of such a study if performed manually by human experts charging at an hourly rate. Even if all human annotations were provided on a volunteer basis, this would still require hundreds if not thousands of hours of cumulative manual human labour. Using our approach, severity annotations for the entire HPO can be generated automatically at a rate of ~100 phenotypes/hour. Further optimisation of the annotation process and increased API rate limits could potentially accelerate this even further.

Throughout this study, we observed that GPT-4 was capable of reliably recovering deep semantic relationships from the medical domain, far beyond making superficial inferences based on lexical similarities. An excellent example of this is the phenotype 'Molar tooth sign on MRI' (`r res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$hpo_id`; severity score=`r round(res_coded$annot_weighted[hpo_name=='Molar tooth sign on MRI']$severity_score_gpt,2)`), which GPT-4 annotated as causing intellectual disability. At first glance, we ourselves assumed this was a false positive as the term appeared to be related to dentition. However, upon further inspection we realised that molar tooth sign is in fact a pattern of abnormal brain morphology that happens to bear some resemblance to molar dentition when observed in radiological scans. This phenotype is a known sign of neurodevelopmental defects that can indeed cause severe intellectual disability [@gleesonMolarToothSign2004].
